Redundant calibration as a complex optimization problem

### 2017-11-17 paper.tex
added most of Jonathan's comments

\begin{abstract}
Observations of the redshifted 21-cm line from the epoch of reionization have recently motivated the construction of low 
frequency radio arrays with highly redundant configurations that provide an alternative calibration strategy - 
''redundant calibration" - and boosts sensitivity on specific spatial scales. In this paper we formulate 
calibration of redundant interferometric arrays as a complex optimization problem \citep[i.e.,][]{Smirnov2015}. 
We then solve this optimization problem via the Levenberg-Marquardt algorithm. This calibration approach is more 
robust to initial conditions than current algorithms and, leveraging on an approximate matrix inversion, allows for further optimization and an efficient implementation (``redundant \textsc{StEfCal}"). We also investigated the use of the preconditioned conjugate gradient method as an alternative optimization strategy, instead of 
an approximate matrix inverse, but found that its computational performance is not competitive with respect to ``redundant \textsc{StEfCal}". The efficient implementation of this new algorithm is made publicly available.
%The construction of new radio interferometers, particularly operating at low frequencies and In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework. We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton. The Gauss-Newton algorithm is currently the de facto standard that is being used to perform redundant calibration. It is common practice to implement redundant calibration via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration. We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method and the Levenberg-Marquardt algorithm with one another.
\end{abstract}

@@

\section{Introduction}
The quest for the redshifted 21-cm line from the Epoch of Reionization (EoR) is a frontier of modern observational cosmology and has motivated the construction of a series of new interferometric arrays operating at low frequencies over the last decade. EoR measurements are challenging because of the intrinsic faintness of the cosmological signal \citep[see, for instance,][for recent reviews]{Furlanetto2016,McQuinn2016} buried underneath foreground emission which is a few orders of magnitude brighter than the EoR anywhere in the sky \citep[e.g.,][]{Bernardi2009,Bernardi2010,Ghosh2012,Dillon2014,Parsons2014}. 
%and that can be separated by leveraging only on the different spectral properties (citations needed!!!). 
As miscalibrated foreground emission leads to artifacts that can jeopardize the EoR signal \citep[e.g.,][]{grobler2014,barry2016,ewall-wice2016}, an exquisite interferometric calibration is required. Such calibration needs an accurate knowledge of both the sky emission and the instrumental response \citep[e.g.,][]{Smirnov2011c}, unless arrays are deployed in a redundant configuration, i.e. with receiving elements placed on regular grids so that they measure many times the same sky brightness emission. Leveraging on this property, redundancy is a promising path to achieve high accurate interferometric calibration \citep[][]{Noordam1982,Wieringa1992,Pearson1984,Liu2010,Noorishad2012,Marthi2014,Sievers2017}. Moreover, redundant configurations provide a sensitivity boost on a selected sample of spatial scales that can be tuned to be the highest signal--to--noise ratio (SNR) EoR modes \citep[][]{Parsons2012,Dillon2016}. These reasons motivated the design and deploment of EoR arrays in redundant configurations like the MIT-EoR \citep{Zheng2014}, the Precision Array to Probe the Epoch of Reionization \citep[PAPER,][]{Ali2015} and the Hydrogen Epoch of Reionization Array \citep[HERA,][]{deboer2017}.

In this paper we present a new algorithm to calibrate redundant arrays based on the complex optimization formalism recently introduced by \cite{Smirnov2015}. 
With respect to current algorithms, it is more robust to initial conditions, while remaining comparatively fast. We also show that given certain approximations this new algorithm reduces to the redundant calibration equivalent of the \textsc{StEfCal} algorithm \citep{Salvini2014}.
%In this paper we will refer to this simplified version of the new algorithm as redundant StEfCal. 
We investigate the speed-up that the preconditioned conjugate gradient (PCG) method provides if it is employed by the new algorithm \citep{Liu2010}.
A comparison between the computational complexity of the optimized new algorithm and redundant StEfCal is also performed.

The paper is organized as follows: we review the complex calibration formalism by \citet{Smirnov2015} in Section~\ref{sec:sky_wirtinger}, we extend it to the redundant case in Section~\ref{sec:red_wirtinger}, we present some computational complexity results in Section~\ref{sec:pcg} and we conclude in Section~\ref{sec:conclusions}.

@@

In a radio interferometer, the true sky visibilities $y_{pq}$ measured by a baseline formed by antenna $p$ and $q$ are always ``corrupted" by the non-ideal response of the receiver, which is often incorporated into a single, receiver-based complex number $g$. The observed visibility $d_{pq}$ is therefore given by \citep{ME1,ME2,RRIME1}
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q} \, y_{pq} + n_{pq},
\end{equation}
where $\conj{(*)}$ indicates conplex conjugation and $n_{pq}$ is the thermal noise component. 
The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a
standard deviation $\sigma$:
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}},
\end{equation}
where $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

@@

Radio interferometric calibration aims to determine the best estimate of $\bg = [g_1,g_2,\cdots,g_N]^T$ in order to correct the data and, following equation~\ref{eq:vis_linear_definition}, can be formulated as a non linear least-squares optimization problem:
\begin{equation}
\label{eq:least_squares}
\min_{\bg} \Lambda(\bg) = \min_{\bg} \|\br\|_F^2 = \min_{\bg} \|\bd - \bv(\bg)\|_F^2, 
\end{equation}
where $\Lambda$ is the objective function, $\br$ is the residual vector and $\|\cdot\|_F$ denotes the Frobenius norm. In standard interferometric calibration, $y_{pq}$ is assumed to be known at some level, for instance through the observation of previously known calibration sources. The knowledge of $y_{pq}$ is often named ``sky model".

The gradient-based minimization algorithms that are generally used to solve non-linear least-squares problems are generally solved by using gradient-based minimization algorithms (i.e. Gauss--Newton -- GN -- or Levenberg--Marquardt -- LM) that require the model ($\bv$ in equation~\ref{eq:least_squares}) to be differentiable towards each parameter. 
When the least squares problem is complex, it becomes less straightforward to apply these gradient-based minimization methods, as
many complex functions are not differentiable if the classic notion of differentiation is used, i.e. $\frac{\partial \conj{z}}{\partial z}$ does not exist if $z \in \mathbb{C}$.

In order to circumvent the differentiability conundrum associated with complex least squares problems, standard interferometric calibration divides the complex optimization problem into its real and imaginary parts and solves for the real and imaginary parts of the unknown model parameters separately. \citet{Smirnov2015} showed, however, that this approach is not needed if complex calculus \citep{Wirtinger1927} is directly adopted. The Wirtinger derivatives are defined as:
\begin{align}
\label{eq:wir}
\frac{\partial}{\partial z} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} -  i \frac{\partial}{\partial y} \right ),&\frac{\partial}{\partial \conj{z}} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} +  i \frac{\partial}{\partial y} \right ), 
\end{align}
which lead to the following relations:
\begin{align}
\label{eq:wir_z}
\frac{\partial z}{\partial z} & = 1, & \frac{\partial \conj{z}}{\partial z}&=0, & \frac{\partial z}{\partial \conj{z}} & = 0, & \frac{\partial \conj{z}}{\partial \conj{z}}&=1.
\end{align}
If the gradient operator is defined using equation~\ref{eq:wir}, the model $\bv$ now becomes analytic in both $\bg$ and $\conj{\bg}$ and equation \ref{eq:wir_z} can be used to derive the complex variants of the real-valued GN and LM algorithms. In the complex GN and LM algorithms, complex parameters and their conjugates are treated as separate variables.

@@

The complex LM update is very similar, the major difference being a damping factor $\lambda$ is introduced:
\begin{equation}
\label{eq:LM_update_skymodel}
\Delta \breve{\bg} = (\bJ^H\bJ + \lambda\bD)^{-1}\bJ^H\breve{\br},
\end{equation}
where $\bD=\bI\odot\bJ^H\bJ$. In this paper, the identity matrix is denoted by $\bI$. Moreover, the Hadamard product and the Hermition transpose are denoted by $\odot$ and $(\cdot)^H$. Note the use of the Wirtinger derivatives in equation~\ref{eq:jac_entries}.
We will refer to $\bJ^H\bJ$ as the Hessian matrix $\bH$ and to $\bJ^H\bJ + \lambda\bD$ as the modified Hessian matrix $\bmH$ throughout this paper.

@@

\citet{Smirnov2015} realized that the diagonal entries of the Hessian matrix $\bH$ are much more significant than its off-diagonal entries, i.e. $\bH$ is nearly diagonal. By approximating $\bH$ by its diagonal and substituting the approximate Hessian matrix the LM parameter update step becomes \citep{Smirnov2015}:
\begin{align}
\breve{\bg}_{k+1} &\approx \frac{1}{1+\lambda}\widetilde{\bH}^{-1}\bJ^H\breve{\bd} + \frac{\lambda}{1+\lambda} \breve{\bg}_k, \nonumber \\
 &= \alpha \widetilde{\bH}^{-1}\bJ^H\breve{\bd} + (1-\alpha)\breve{\bg}_k, 
\label{eq:stef_alpha}  
\end{align}
where $\alpha = \frac{1}{1+\lambda}$. Note that equation~\ref{eq:one_half} and equation~\ref{eq:stef_alpha} are   
not dependent on $\breve{\br}$.

Interestingly enough, if $\lambda = 0$ we obtain the odd parameter update step of \textsc{StEfCal}\footnote{$k\in\{0,2,\cdots\}$}, and if $\lambda=1$ (which corresponds
to $\alpha=\frac{1}{2}$) we obtain the even parameter update step of \textsc{StEfCal}\footnote{$k\in\{1,3,\cdots\}$} \citep[\textsc{StEfCal},][]{Mitchell:MWA-cal,Salvini2014}. In the \textsc{StEfCal} algorithm, the measurement equation, equation~\ref{eq:vis_definition}, is linearized by assuming that the gains are known, but that their conjugates are not. Under this assumption, the system of equations become linear and the conjugates of the gains can be obtained straightforwardly. Starting from the latest value of the gain conjugates, an updated estimate of the gains can be obtained and the succession continues until convergence is reached.
Alternating between solving and fixing different sets of parameters (which is exactly what \textsc{StEfCal} does) is in general referred to 
as the alternating direction implicit method (ADI). The \textsc{StEfCal} algorithm reduces the computational complexity of calibration from $O(N^3)$ to $O(N^2)$.

\section{Redundant Wirtinger Calibration}
\label{sec:red_wirtinger}
Interferometric baselines are named redundant when they sample the exact same visibilities in the $uv$-plane, i.e. if baseline $pq$ and $rs$ are redundant then $y_{pq} = y_{rs}$. It is convenient to group redundant visibilities together and label each group using a single index rather than using their antenna pairs as in equation~\ref{eq:vis_definition}. We introduce a function $\phi_{pq}$ that maps the antenna pair associated with a specific baseline to the single group index of the redundant baseline group that includes the baseline in question, i.e. if baseline $pq$ and $rs$ are redundant then $\phi_{pq} = \phi_{rs}$. The function $\phi_{pq}$ is not symmetric as $\phi_{pq} = 0$ if $p \geq q$. Equation~\ref{eq:vis_definition} can be re-written for a redundant array as:
\begin{equation}
\label{eq:vis_red}
d_{pq} = g_{p}\conj{g_q}y_{\phi_{pq}} + n_{pq},
\end{equation}
with the same vector form as equation~\ref{eq:vis_linear_definition} if
\begin{align}
 \left [ \bd \right]_{\alpha_{pq}} &= d_{pq}, & \left [ \bv \right ]_{\alpha_{pq}} &= v_{pq}=g_p y_{\phi_{pq}} \conj{g_q},\nonumber\\
 \left [ \bn \right ]_{\alpha_{pq}} &= n_{pq}, &  &\label{eq:vec_definitions}
\end{align}
where the vectors in Equation~\ref{eq:vec_definitions} are column vectors of size $B$ (i.e. $p<q$).

We also introduce the following symmetric variant of $\phi_{pq}$:
\begin{equation}
\zeta_{pq} = 
\begin{cases}
\phi_{pq}~\textrm{if}~p \leq q\\
\phi_{qp}~\textrm{if}~p>q
\end{cases},
\end{equation} 
and we will refer to $\zeta_{pq}$ as the symmetric geometric function.
It is possible to construct a simple analytic expression for $\zeta_{pq}$ for an east-west regular array, i.e. $\zeta_{pq} = |q-p|$. It becomes, however, increasingly difficult to construct analytic expressions of $\zeta_{pq}$ for more complicated array layouts. The empirically constructed symmetric geometry functions of three different redundant layouts are displayed in Figure~\ref{fig:geometry_function}. We denote the range of $\zeta_{pq}$ with $\mathcal{R}(\zeta_{pq})$. The maximal element that $\zeta_{pq}$ can ascertain is denoted by $L$ and can be interpreted as the maximal number of unique redundant baseline groups which can be formed for a given array layout.

@@

\subfigure[Hexagonal: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./HEX_phi.pdf}\label{fig:HEX_phi}}
\subfigure[Square: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./SQR_phi.pdf}\label{fig:SQR_phi}}
\subfigure[Regular east-west: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./REG_phi.pdf}\label{fig:REG_phi}}
\caption{Three different redundant antenna layouts: hexagonal (top left), square (top middle) and regular east-west (top right) with their respective symmetric redundancy geometry functions $\zeta_{pq}$ (bottom panels). We used 91 antennas to construct the hexagonal layout, while 100 antennas were used in the square and east-west layouts. In the case of the east-west layout we only plot the positions of the first ten antennas. The maximal amount of redundant baseline groups $L$ that can be formed for 
the hexagonal, square and east-west layouts are 165, 180 and 99 respectively. The analytic expressions of $L$ for an hexagonal, square and east-west layout is $L = 2N-\frac{1}{2}\sqrt{12N-3}-\frac{1}{2}$,
$L=2N-2\sqrt{N}$ and $L=N-1$.\label{fig:geometry_function}}
\end{figure*}

@@

The simulated visibilities included a noise term. The purpose of which is to produce a certain SNR in the simulated
visibilities. We made use of the following definition of \citep[SNR, ][]{Liu2010,Marthi2014}:  
\begin{equation}
\label{}
\textrm{SNR} = 10\log \left (\frac{<\bv\odot\conj{\bv}>_{\nu,t,pq}}{<\bn\odot\conj{\bn}>_{\nu,t,pq}} \right ), 
\end{equation}
where $<*>_{\nu,t,pq}$ denotes averaging over frequency, time and baseline.

Figure~\ref{fig:prec_error} show the simulation results as a function of SNR and number of antennas. The accuracy of our solutions is quantified through the percentage error
\begin{equation}
\label{eq:beta}
\beta = \frac{\|\bv - \widehat{\bv}\|_F^2}{\|\bv\|_F^2},
\end{equation}
where $\widehat{\bv}$ is the redundant \textsc{StEfCal} parameter estimate.

@@

\begin{figure}
\includegraphics[width=0.47\textwidth]{./prec_error.pdf} 
\caption{We plot the percentage error $\beta$ between the simulated visibilitied and the visibilities solved for by redundant \textsc{StEfCal} for different SNR values as a function of the number of antennas in the array $N$.}
\label{fig:prec_error}
\end{figure}

>>>>

\begin{abstract}
Observations of the redshifted 21-cm line from the epoch of reionization have recently motivated the construction of low 
frequency radio arrays with highly redundant configurations. These configurations provide an alternative calibration strategy - 
''redundant calibration" - and boosts sensitivity on specific spatial scales. In this paper, we formulate 
calibration of redundant interferometric arrays as a complex optimization problem {\bf SJK: suggests removing citation} \citep[i.e.,][]{Smirnov2015}. 
We solve this optimization problem via the Levenberg-Marquardt algorithm. This calibration approach is more 
robust to initial conditions than current algorithms and, by leveraging an approximate matrix inversion, allows for further optimization and an efficient implementation (``redundant \textsc{StEfCal}"). {\bf (SJK: ...boasts competitive [big O notation here] scaling. We coin the term "redundant StEfCal" for our algorithm and implementation)}.  We also investigated using the preconditioned conjugate gradient method as an alternative to the approximate matrix inverse, but found that its computational performance is not competitive with respect to ``redundant \textsc{StEfCal}". The efficient implementation of this new algorithm is made publicly available.
%The construction of new radio interferometers, particularly operating at low frequencies and In this paper we show that instead of splitting our redundant calibration problem into its phase-and-amplitude components (the more traditional approach) we can treat the complex unknown model parameters and their complex conjugates as separate unknowns. We also derive the Levenberg-Maquardt and Gauss-Newton update steps associated with this alternative parameter-and-conjugate framework. We focus more on the Levenberg-Maquardt algorithm in this paper as it exhibits better convergence properties when compared to Gauss-Newton. The Gauss-Newton algorithm is currently the de facto standard that is being used to perform redundant calibration. It is common practice to implement redundant calibration via a hierarchical algorithmic strategy. Using the Levenberg-Maquardt algorithm, therefore, reduces the number of algorithmic layers that are needed to perform redundant calibration. We also compare two previously proposed methods to accelerate redundant calibration, namely the alternating direction implicit method and the preconditioned conjugate gradient method. We found that the alternating direction implicit method outperforms the preconditioned conjugate gradient method. However, the preconditioned conjugate gradient method is more robust as it is more straightforward to adapt to other calibration use cases. We make use of the parameter-and-conjugate framework in this paper, rather than the more traditional amplitude-and-phase approach as it makes it trivial to relate the alternating direction implicit method and the Levenberg-Marquardt algorithm with one another.
\end{abstract}

@@

\section{Introduction}
The quest for the redshifted 21-cm line from the Epoch of Reionization (EoR) is a frontier of modern observational cosmology and has motivated the construction of a series of new interferometric arrays operating at low frequencies over the last decade. EoR measurements are challenging because of the intrinsic faintness of the cosmological signal \citep[see, for instance,][for recent reviews]{Furlanetto2016,McQuinn2016} buried underneath foreground emission which is a few orders of magnitude brighter than the EoR anywhere in the sky \citep[e.g.,][]{Bernardi2009,Bernardi2010,Ghosh2012,Dillon2014,Parsons2014}. 
%and that can be separated by leveraging only on the different spectral properties (citations needed!!!). 
As miscalibrated foreground emission leads to artifacts that can jeopardize the EoR signal \citep[e.g.,][]{grobler2014,barry2016,ewall-wice2016}, an exceptional interferometric calibration is required. Such calibration needs an accurate knowledge of both the sky emission and the instrumental response \citep[e.g.,][]{Smirnov2011c}, unless arrays are deployed in a redundant configuration, i.e. with receiving elements placed on regular grids so that multiple baselines measure the same sky brightness emission. Leveraging this property, redundancy is a promising path to achieve highly accurate interferometric calibration \citep[][]{Noordam1982,Wieringa1992,Pearson1984,Liu2010,Noorishad2012,Marthi2014,Sievers2017}. Moreover, redundant configurations provide a sensitivity boost on a selected sample of spatial scales that can be tuned to be the highest signal--to--noise ratio (SNR) EoR modes \citep[][]{Parsons2012,Dillon2016}. These reasons motivated the design and deploment of EoR arrays in redundant configurations like the MIT-EoR \citep{Zheng2014}, the Precision Array to Probe the Epoch of Reionization \citep[PAPER,][]{Ali2015} and the Hydrogen Epoch of Reionization Array \citep[HERA,][]{deboer2017}.

In this paper we, present a new algorithm to calibrate redundant arrays based on the complex optimization formalism recently introduced by \cite{Smirnov2015}. 
With respect to current algorithms, it is more robust to initial conditions, while remaining comparatively fast. We also show that given certain approximations this new algorithm reduces to the redundant calibration equivalent of the \textsc{StEfCal} algorithm \citep{Salvini2014}.
%In this paper we will refer to this simplified version of the new algorithm as redundant StEfCal. 
We investigate the speed-up that the preconditioned conjugate gradient (PCG) method provides if it is employed by the new algorithm \citep{Liu2010}.
A comparison between the computational complexity of the optimized new algorithm and redundant StEfCal is also performed.

The paper is organized as follows: we review the complex calibration formalism by \citet{Smirnov2015} in Section~\ref{sec:sky_wirtinger}, we extend it to the redundant case in Section~\ref{sec:red_wirtinger}, we present some computational complexity results in Section~\ref{sec:pcg} and we presnt our conclusions in Section~\ref{sec:conclusions}.

@@

In a radio interferometer, the true sky visibilities $y_{pq}$ measured by a baseline formed by antenna $p$ and $q$ are always ``corrupted" by the non-ideal response of the receiver, which is often incorporated into a single, receiver-based complex number $g$ (i.e. an antenna gain). The observed visibility $d_{pq}$ is therefore given by \citep{ME1,ME2,RRIME1}
\begin{equation}
\label{eq:vis_definition}
d_{pq} = g_{p}\conj{g_q} \, y_{pq} + n_{pq},
\end{equation}
where $\conj{(*)}$ indicates conplex conjugation and $n_{pq}$ is the thermal noise component. 
The real and imaginary components of the thermal noise are normally distributed with a mean of zero and a
standard deviation $\sigma$:
\begin{equation}
\sigma \propto \frac{T_{\textrm{sys}}}{\sqrt{\Delta \nu \tau}},
\end{equation}
where $T_{\textrm{sys}}$ is equal to the system temperature, $\Delta \nu$ is the observational bandwidth and $\tau$ is the integration time per visibility.

@@

{\bf JSK: I would introduce the alpha quantity on its own before attaching it as an index in the definition of d. The reason for this is that eq. 4 and 5 seem very complicated without first understanding the meaning of alpha. Could have a paragraph starting ``In order to express equation 1 in vector form, we first introduce $\alpha$ which maps...''}

{\bf TLG: Was like this earlier, got changed during the edits. GB let me know if it should be changed back.}

Radio interferometric calibration aims to determine the best estimate of $\bg = [g_1,g_2,\cdots,g_N]^T$ in order to correct the data and, following equation~\ref{eq:vis_linear_definition}, can be formulated as a non linear least-squares optimization problem:
\begin{equation}
\label{eq:least_squares}
\min_{\bg} \Lambda(\bg) = \min_{\bg} \|\br\|_F^2 = \min_{\bg} \|\bd - \bv(\bg)\|_F^2, 
\end{equation}
where $\Lambda$ is the objective function, $\br$ is the residual vector and $\|\cdot\|_F$ denotes the Frobenius norm. In standard interferometric calibration, $y_{pq}$ is assumed to be known at some level, for instance through the observation of previously known calibration sources.

Non-linear least-squares problems are generally solved by using gradient-based minimization algorithms (i.e. Gauss--Newton (GN) -- or Levenberg--Marquardt (LM)) that require the model ($\bv$ in equation~\ref{eq:least_squares}) to be differentiable towards each parameter. 
When the least squares problem is complex, it becomes less straightforward to apply these gradient-based minimization methods, as
many complex functions are not differentiable if the classic notion of differentiation is used, i.e. $\frac{\partial \conj{z}}{\partial z}$ does not exist if $z \in \mathbb{C}$.

In order to circumvent the differentiability conundrum associated with complex least squares problems, standard interferometric calibration divides the complex optimization problem into its real and imaginary parts and solves for the real and imaginary parts of the unknown model parameters separately. \citet{Smirnov2015} showed, however, that this approach is not needed if complex calculus \citep{Wirtinger1927} is directly adopted {\bf JSK: employed}. The Wirtinger derivatives are defined as:
\begin{align}
\label{eq:wir}
\frac{\partial}{\partial z} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} -  i \frac{\partial}{\partial y} \right ),&\frac{\partial}{\partial \conj{z}} &= \frac{1}{2}\left ( \frac{\partial}{\partial x} +  i \frac{\partial}{\partial y} \right ), 
\end{align}
which lead to the following relations:
\begin{align}
\label{eq:wir_z}
\frac{\partial z}{\partial z} & = 1, & \frac{\partial \conj{z}}{\partial z}&=0, & \frac{\partial z}{\partial \conj{z}} & = 0, & \frac{\partial \conj{z}}{\partial \conj{z}}&=1.
\end{align}
If the gradient operator is defined using equation~\ref{eq:wir}, the model $\bv$ now becomes analytic in both $\bg$ and $\conj{\bg}$ and equation \ref{eq:wir_z} can be used to derive the complex variants of the real-valued GN and LM algorithms. In the complex GN and LM algorithms, complex parameters and their conjugates are treated as separate variables.

@@

The complex LM update is very similar, the major difference being the introduction of a damping parameter, $\lambda$:
\begin{equation}
\label{eq:LM_update_skymodel}
\Delta \breve{\bg} = (\bJ^H\bJ + \lambda\bD)^{-1}\bJ^H\breve{\br},
\end{equation}
where $\bD=\bI\odot\bJ^H\bJ$. In this paper, the identity matrix is denoted by $\bI$. Moreover, the Hadamard product and the Hermition transpose are denoted by $\odot$ and $(\cdot)^H$ respectively. Note the use of the Wirtinger derivatives in equation~\ref{eq:jac_entries}.
We will refer to $\bJ^H\bJ$ as the Hessian matrix $\bH$ and to $\bJ^H\bJ + \lambda\bD$ as the modified Hessian matrix $\bmH$ throughout this paper.

@@

\citet{Smirnov2015} realized that the diagonal entries of the Hessian matrix $\bH$ are much more significant than its off-diagonal entries, i.e. $\bH$ is nearly diagonal. By approximating $\bH$ by its diagonal and substituting the approximate Hessian matrix the LM parameter update step becomes \citep{Smirnov2015}:
\begin{align}
\breve{\bg}_{k+1} &\approx \frac{1}{1+\lambda}\widetilde{\bH}^{-1}\bJ^H\breve{\bd} + \frac{\lambda}{1+\lambda} \breve{\bg}_k, \nonumber \\
 &= \rho \widetilde{\bH}^{-1}\bJ^H\breve{\bd} + (1-\rho)\breve{\bg}_k, 
\label{eq:stef_alpha}  
\end{align}
where $\rho = \frac{1}{1+\lambda}$. Note that equation~\ref{eq:one_half} and equation~\ref{eq:stef_alpha} are   
not dependent on $\breve{\br}$.

Interestingly enough, if $\lambda = 0$ we obtain the odd parameter update step of \textsc{StEfCal}\footnote{$k\in\{0,2,\cdots\}$}, and if $\lambda=1$ (which corresponds
to $\rho=\frac{1}{2}$) we obtain the even parameter update step of \textsc{StEfCal}\footnote{$k\in\{1,3,\cdots\}$} \citep[\textsc{StEfCal},][]{Mitchell:MWA-cal,Salvini2014}. In the \textsc{StEfCal} algorithm, the measurement equation (equation~\ref{eq:vis_definition}) is linearized by assuming that the gains are known, but that their conjugates are not. Under this assumption, the system of equations become linear and the conjugates of the gains can be obtained in a straightforward manner. Starting from the latest value of the gain conjugates, an updated estimate of the gains can be obtained iteratively until convergence is reached.
Alternating between solving and fixing different sets of parameters (which is exactly what \textsc{StEfCal} does) is referred to 
as the alternating direction implicit (ADI) method. The \textsc{StEfCal} algorithm reduces the computational complexity of calibration from $O(N^3)$ to $O(N^2)$.

\section{Redundant Wirtinger Calibration}
\label{sec:red_wirtinger}
Interferometric baselines are redundant when they sample the exact same visibilities in the $uv$-plane, i.e. if baseline $pq$ and $rs$ are redundant then $y_{pq} = y_{rs}$. 
A redundant array layout allows us to solve for the unknown observed visibilities themselves in
addition to the antenna gains (see equation \ref{eq:vec_linear_definitions}). This is true, since in the case of a redundant layout equation \ref{eq:vec_linear_definitions} is already an overdetermined 
system (there is no need to first predict visibilities from a pre-existing skymodel).

{\bf SJK: I feel there needs to be some additional emphasis placed on the fact that the model visibilities are a solvable parameter. I found a great deal of emphasis is placed on the redundancy, but not why that redundancy is so crucial to the solution procedure.}
{\bf TLG: Added paragraph, above, GB want to phrase it better??}

%When we employ a redundant array, we dramatically reduce the number 
%of unknowns (assuming we have no prior knowledge of the sky). If our array is redundant enough we reduce
%the number of unknowns to such an extent that we are able
%to solve for  Redundant calibration is the
%procedure by which we estimate both the unknown observed
%visibilities and the antenna gains.

It is convenient to group redundant visibilities together and label each group using a single index rather than using their antenna pairs as in equation~\ref{eq:vis_definition}. We introduce a function $\phi$ that maps the antenna pair associated with a specific baseline to its corresponding redundant baseline group, i.e. if baseline $pq$ and $rs$ are redundant then $\phi_{pq} = \phi_{rs}$ (implying they belong to the same group). The function $\phi_{pq}$ is not symmetric as $\phi_{pq} = 0$ if $p \geq q$. Equation~\ref{eq:vis_definition} can be re-written for a redundant array as:
\begin{equation}
\label{eq:vis_red}
d_{pq} = g_{p}\conj{g_q}y_{\phi_{pq}} + n_{pq},
\end{equation}
with the same vector form as equation~\ref{eq:vis_linear_definition} if
\begin{align}
 \left [ \bd \right]_{\alpha_{pq}} &= d_{pq}, & \left [ \bv \right ]_{\alpha_{pq}} &= v_{pq}=g_p y_{\phi_{pq}} \conj{g_q},\nonumber\\
 \left [ \bn \right ]_{\alpha_{pq}} &= n_{pq}, &  &\label{eq:vec_definitions}
\end{align}
where the vectors in Equation~\ref{eq:vec_definitions} are column vectors of size $B$ (i.e. $p<q$).

We also introduce the following symmetric variant of $\phi_{pq}$:
\begin{equation}
\zeta_{pq} = 
\begin{cases}
\phi_{pq}~\textrm{if}~p \leq q\\
\phi_{qp}~\textrm{if}~p>q
\end{cases},
\end{equation} 
and we will refer to $\zeta_{pq}$ as the symmetric geometric function.
It is possible to construct a simple analytic expression for $\zeta_{pq}$ for an east-west regular array, i.e. $\zeta_{pq} = |q-p|$. It becomes, however, increasingly difficult to construct analytic expressions of $\zeta_{pq}$ for more complicated array layouts. The empirically constructed symmetric geometry functions of three different redundant layouts are displayed in Figure~\ref{fig:geometry_function}. We denote the range of $\zeta_{pq}$ with $\mathcal{R}(\zeta_{pq})$. The maximal element that $\zeta_{pq}$ can ascertain is denoted by $L$ and can be interpreted as the greatest number of unique redundant baseline groups which can be formed for a given array layout.

@@

\subfigure[Hexagonal: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./HEX_phi.pdf}\label{fig:HEX_phi}}
\subfigure[Square: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./SQR_phi.pdf}\label{fig:SQR_phi}}
\subfigure[Regular east-west: $\zeta_{pq}$]
{\includegraphics[width=0.33\textwidth]{./REG_phi.pdf}\label{fig:REG_phi}}
\caption{Three different redundant antenna layouts: hexagonal (top left), square (top middle) and regular east-west (top right) with their associated symmetric redundancy geometry functions $\zeta_{pq}$ (bottom panels). We used 91 antennas to construct the hexagonal layout, while 100 antennas were used in the square and east-west layouts. In the case of the east-west layout we only plot the positions of the first ten antennas. The maximal amount of redundant baseline groups $L$ that can be formed for 
the hexagonal, square and east-west layouts are 165, 180 and 99 respectively. The analytic expressions of $L$ for an hexagonal, square and east-west layout are $L = 2N-\frac{1}{2}\sqrt{12N-3}-\frac{1}{2}$,
$L=2N-2\sqrt{N}$ and $L=N-1$ respectively.\label{fig:geometry_function}}
\end{figure*}

@@

The simulated visibilities included a noise term, the purpose of which was to produce a certain SNR in the simulated
visibilities. We made use of the following definition of \citep[SNR, ][]{Liu2010,Marthi2014}:  
\begin{equation}
\label{}
\textrm{SNR} = 10\log \left (\frac{<\bv\odot\conj{\bv}>_{\nu,t,pq}}{<\bn\odot\conj{\bn}>_{\nu,t,pq}} \right ), 
\end{equation}
where $<*>_{\nu,t,pq}$ denotes averaging over frequency, time and baseline.

Figure~\ref{fig:prec_error} shows the simulation results as a function of SNR and number of antennas. The accuracy of our solutions is quantified through the percentage error
\begin{equation}
\label{eq:beta}
\beta = \frac{\|\bv - \widehat{\bv}\|_F^2}{\|\bv\|_F^2},
\end{equation}
where $\widehat{\bv}$ is the redundant \textsc{StEfCal} parameter estimate.

@@

\begin{figure}
\includegraphics[width=0.47\textwidth]{./prec_error.pdf} 
\caption{We plot the percentage error $\beta$ between the simulated visibilities and the visibilities solved for by redundant \textsc{StEfCal} for different SNR values as a function of the number of antennas ($N$) in the array.}
\label{fig:prec_error}
\end{figure}

\section{Preconditioned Conjugate Gradient Method}
\label{sec:pcg}
\citet{Liu2010} suggested that the speed of convergence of redundant calibration could be optimized using the 
conjugate gradient method \citep{Hestenes1952}, which would be computationally advantageous, since the Hessian matrix associated 
with redundant calibration (see Figure~\ref{fig:hessian}) is sparse \citep{Reid1971}. In this section we study the computational complexity
of the conjugate gradient (CG) method when it is used to invert the modified Hessian matrix (see equation~\ref{eq:LM_update}), 
in particular when preconditioning is used (i.e. the PCG method). Interestingly, empirical tests suggest that the unmodified Hessian itself is singular. It is therefore important to mention, that the CG method 
can pseudo invert the unmodified Hessian as well, i.e. the CG method can be directly applied to equation~\ref{eq:GN_update}, because the Hessian is a positive semi-definite Hermitian matrix and the vector $\bJ^H\breve{\br}$ is an element of its column range \citep{Lu2015}.
